Modeling memory and taxis to analyze migrations

### 2021-05-28 text/ms.Rmd
More ms

Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].  For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996]. The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2013].

@@

where  $u(x,t)$ represents the population distributed in time and space, $h(x,t)$ is the distribution of the resource, which in dynamic and patchy as detailed below, $\varepsilon$ is the diffusion rate and$\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), $\beta$ is the strength of advection up the memory gradient.  Finally, there is a  memory based speed term $v_m(x,t,y)$, described below.

The population has a "social memory" of the mean location and speed at which the population moved exactly one year prior, i.e. $\widehat{u}(t-\tau)$, where $\tau$ is the duration of a single year and the $\hat{\,}$ operator signifies the expected value of the population distribution. This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:

$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t - y\tau) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)\,\,\,\,(2)$$

Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e. $y = \lfloor t/\tau \rfloor$ and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory. When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.  Importantly, the individuals remember the movement rate of their mean location:  $d\widehat{u_m}(t)/dt$.

The overarching behavior of the population is to attempt to follow the memory.  Thus the speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e. for those animals that are lagging behind the optimal remembered location, the population's speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.

The specific relationship is essentially a logistic curve:

@@

The model is confined to a one-dimensional bounded domain $[0,\chi]$, with no flux outside of the boundaries i.e. ${\frac{\partial u(0,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$,  Thus the total population is fixed at all times and found within the domain.

Note, this model does not contain any explicit birth or death processes, the population is fixed, and the parameters remain constant throughout time.  Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic environment.

@@

## Environment and seasonal resource

We ran our numerical experiments on a spatial domain ranging from 0 to $\chi = 100$, and a period of duration $\tau = 100$ (i.e.~a 100 day "year").  We are interested in an approximately periodic resource dynamic, i.e. one in which $h(x,t \approx h(x, t-\tau))$ and $\tau$ is the period of the process. Furthermore, we were mainly interested in a "non-surfable" resource where the patches are isolated and pulse periodically.

Thus, the resource model consists of two bivariate normal pulses centered at $\mu_p$ at $t_p$ and $\chi - \mu_p$ at $\tau - t_p$ (e.g. location 80 and 20 at times 25 and 75, respectively). These pulses have a shared time scale of duration $\sigma_t$ and a spatial scale of extent $\sigma_x$, the standard deviation in the time and space dimension respectively.  This flexible formulation allows us to compare long (duration) and narrow (spatially), long and wide, short and narrow, and short and wide pulses, and to easily simulate scenarios where there is drift or randomness in the location or timing of the pulses. Note that in order to make the magnitude of the resource pulse comparable to the magnitude of the memory, which integrates to 1 at every time point, we normalize the resource such that its average value across the year is also 1.

@@

To quantify the foraging efficiency, i.e., the organisms’ ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya Coefficient [@Bhattacharyya1943] which quantifies the similarity between two distributions.  We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE). Thus, the foraging efficiency index is: 
$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
where the spatial integral is taken over the domain.  This metric is constraned to be between 0 and 1.

Simulations are run until a (quasi)-equilibrium state is achieved, i.e. where the Bhattacharya index across subseqyuent years reaches a value of 0.9999.  Once that state is attained, we compute the mirgatoriness and foraging efficiency metrics, and additionally collect the number of years that were required to reach stationarity.

>>>>

Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].  For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996]. The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2013].

@@

where  $u(x,t)$ represents the population distributed in time and space, $\varepsilon$ is the diffusion rate and $h(x,t)$ is the distribution of the resource.  The resource is structured and seasonal with period $\tau$, such that $h(x, t) \approx h(x, t-\tau)$. $\alpha$ represents the strength of advection in the direction of the resource $h(x,t)$ gradient, as standard in chemotaxis equations.  Finally, there is the social memory based speed term $v_m(x,t,y)$, which evolves with the year $y \in \{0,1,2,...\}$, i.e. the count of periods $\tau$: $y = \lfloor t/\tau \rfloor$.

The social memory term works as follows:  The population retains a memory of the mean location of the population exactly one year prior, i.e. $\widehat{u}(t-\tau)$, where $\hat{\,}$ operator signifies the expected value of the population distribution. This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:

$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)\,\,\,\,(2)$$

The long-term, reference memory term, which does not change throughout the evolution of the process, can be thought of a strictly periodic function.  The parameter $\kappa \in (0,1)$ is the relative contribution of that long-term memory. When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.  Importantly, the individuals remember not just the location of the actionable memory term, but the movement rate of that term as well, i.e. $d\widehat{u_m}(t)/dt$.

The overarching behavior of the population is to attempt to repeat that memory.  Thus the memory speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e. for those animals that are lagging behind the optimal remembered location, the memory speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.

The specific relationship is a logistic curve:

@@

The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e. ${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$.  Thus the total population is fixed at all times - there are no birth or death processes - and found within the domain.   Furthermore, the parameters remain constant throughout time.  Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic and changing environment.

@@

## Seasonal resource

We ran our numerical experiments on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day years).  We were interested in an approximately periodic resource dynamic, i.e. one in which $h(x,t \approx h(x, t-\tau))$. We generated two types of resource distributions. A "non-surfable" resource (*island resource*), and weakly surfable resource (*drifting resource*).  Both are characterized by a peak in time and space centered at $M_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 40 and -40 at times 25 and 75, respectively). These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively.  The island resource is simply two uncorrelated bivariate normal distributions $$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$
where the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e. ${1 \over \tau} \int_T\int_X h(x,t) dx\,dt = 1$.

The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1.  This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).

Within a given year, the resource is entirely symmetric: $h(x,t) = h(-x, \tau-t)$.  However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to:  $m_x(y) \sim {\cal N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {\cal N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.  Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.  These trends mirror the poleward shift of peak resources and the earlier spring phenology occurring with a warming global climate.  In the scenarios we explore, the spatial scales ($s_x$ and $s_t$) remain constant.

@@

To quantify the foraging efficiency, i.e., the organisms’ ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya Coefficient [@Bhattacharyya1943] which quantifies the similarity between two distributions.  We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE). Thus, the foraging efficiency index is: 
$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
where the spatial integral is taken over the domain.  This metric is constrained to be between 0 and 1.

For simulations with a constant resource, we run those until a quasi-equilibrium state is achieved, i.e. where the Bhattacharya index of the population distribution across subsequent years reaches a value of 0.9999.  Once that state is attained, we compute the mirgatoriness and foraging efficiency metrics, and additionally collect the number of years that were required to reach stationarity.

For numerical runs with climate change, we run the simulation for 30 years, since equilibrium is never achieved) and retain the foraging efficiency for the final 10 years, reflecting something like the dynamic long-term average of the system's ability to keep track of climate change.

We also record the actual memory.
